Tail associations and their impact on extinction risk

### 2019-10-29 MainText.Rmd
edited notations

```{r read_res, echo=F}
ktau_common<-readRDS("./Results/ktau_common.RDS")
```

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Kendall's $\tau$ again $0.75$, up to sampling variation), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\tau=0.75$ between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variation), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cummulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Kendall correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$. The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with (at $t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and moderate to extreme left 
tail- and right-tail associated environmental noises with same Kendall's $\tau$ (= `r ktau_common`)
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got a similar trend for extinction risks with the different tail associations (left and right).
The non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{fig_extrisk_agestr} represents this scenario when vital rates have moderate to extreme tail association in either tail but with same Kendall's correlation ($\tau =$ `r ktau_common`). Extinction risks for two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{fig_extrisk_agestr}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr}C or D). This result answered (Q1) and showed extinction risk of 
a population could be considered as a direct manifestation of tail association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered for each patch; last three columns indicate equilibrium population value ($P_e^{i}$), critical growth rate ($r_c^{i}$ : transition point from under- to over-compensatory dynamics of deterministic models) and bifurcation parameter ($r_{bf}^{i}$ : transition point from stable to unstable equilibrium of deterministic models), respectively. Superscript $i=1,2,...,n$ indicates index for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
<!--***DAN: Shyamolina, in this table and in the sup mat (and anywhere else where this comment applies),
please replace the notation P^i_t and \epsilon^i_t by P_i(t) and \epsilon_i(t), the reason being,
putting an i in the superscript makes it look like and exponent. This is legitimately confusing because
we use lots of other exponents. I want you to generally refrain from placing the habitat patch index in the
superscript, putting in the subscript instead. This means you have to put the t in parens in front of the
variable in question, instead of in the subscript. This will require several changes to this table. Please
make those changes and then commit. Then please make the corresponding changes to the sup mat, whereever needed.-->
<!--***DAN: Dan, once Shyamolina does the above, you will then have to go back and edit this table in various
ways that you cannot do until after that change is made.-->

>>>>

```{r read_res, echo=F}
srho_common<-readRDS("./Results/srho_common.RDS")
```

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variation), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variation), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cummulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$. The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with (at $t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and moderate to extreme left 
tail- and right-tail associated environmental noises with same Kendall's $\rho$ (= `r srho_common`)
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got a similar trend for extinction risks with the different tail associations (left and right).
The non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{fig_extrisk_agestr} represents this scenario when vital rates have moderate to extreme tail association in either tail but with same Kendall's correlation ($\rho =$ `r srho_common`). Extinction risks for two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{fig_extrisk_agestr}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr}C or D). This result answered (Q1) and showed extinction risk of 
a population could be considered as a direct manifestation of tail association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered for each patch; last three columns indicate equilibrium population value ($P_{e}$), critical growth rate ($r_{c}$ : transition point from under- to over-compensatory dynamics of deterministic models) and bifurcation parameter ($r_{bf}$ : transition point from stable to unstable equilibrium of deterministic models), respectively. Subscript $i=1,2,...,n$ indicates index for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
<!--***DAN: Shyamolina, in this table and in the sup mat (and anywhere else where this comment applies),
please replace the notation P^i_t and \epsilon^i_t by P_i(t) and \epsilon_i(t), the reason being,
putting an i in the superscript makes it look like and exponent. This is legitimately confusing because
we use lots of other exponents. I want you to generally refrain from placing the habitat patch index in the
superscript, putting in the subscript instead. This means you have to put the t in parens in front of the
variable in question, instead of in the subscript. This will require several changes to this table. Please
make those changes and then commit. Then please make the corresponding changes to the sup mat, whereever needed.-->
<!--***DAN: Dan, once Shyamolina does the above, you will then have to go back and edit this table in various
ways that you cannot do until after that change is made.-->
